Explorative data structure comparison with application to \ldots

### 2017-03-01 article/article_v0.01.tex
opdateret manuskript
opdateret manuskript

\usepackage[myheadings]{fullpage}
\usepackage{pmetrika}
\usepackage{pmbib}
\usepackage{color}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{hyperref}

@@

\author{Anne H. Petersen, Bo Bang, Karl Markussen}

@@

\section{Introduction}
\label{sec:introduction}
The origin stories of datasets have changed over time. While in the past, data was often collected with a specific scientific inquiry in mind, today, a lot of data is accumulated without such a specific purpose. This is due to vast amounts of data being registered online and due to a trend towards more open source research. The latter phenomenon in particular poses new challenges wrt. data quality assessment. When data are collected and made public without a specific end-point in mind, how do we ensure that e.g.\ differences in measurement instruments do not cause the data to be effectively divided into subsets that are simply not comparable?

@@

\subsection{Principal component analysis - a very brief introduction}

{\color{red}

\emph{Principal component analysis} (PCA) may be interpreted in terms of the \emph{rank-q-reconstruction error}. To describe this consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, and let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ be their average. Suppose that we want to describe each of the observations by $q$ numbers instead of the original $d$ numbers. The rank-q-reconstruction error is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*} 
PCA ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of the empirical covariance $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different. If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{j=q+1}^d \lambda_j$. In the terminology of PCA the eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}. The $j$th loading can also found as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}

It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and associated variation components $\lambda_j$ describe the structure of the dataset simultaneously for all approximating dimension $q$. This implies that the loadings and variation components can be used to investigate the structure of the dataset without the need to decide on an approximating dimension $q$.
}

@@

\vspace{\fill}\clearpage

\begin{thebibliography}

\bibitem Ackerman, T.A. (1992). \hl{EXAMPLE POST.} A didactic explanation of item bias, item impact, and item validity from a multidimensional perspective. \textit{Journal of Educational Measurement}, \textit{29}, 67--91.

\end{thebibliography}
\vspace{\fill}

%\vfill\eject
\end{document}

>>>>

\usepackage[myheadings]{fullpage}
\usepackage{pmetrika}
\usepackage{pmbib}
\usepackage{color}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{natbib}

@@

\author{Anne H. Petersen, Bo Markussen, Karl Bang Christensen}

@@

\section{Introduction}
\label{sec:introduction}

Classical statistical methodology is aimed at analyzing data from designed experiments and historically statistical analyses have been done by researchers who knew the design and origin story of the data set well. The origin stories of data sets have changed over time and today a lot of data is accumulated without specific purpose. This is due to vast amounts of data being registered online and to a trend towards more open source research. The latter phenomenon in particular poses new challenges wrt. data quality assessment. When data are collected and made public without a specific end-point in mind, how do we ensure that differences in, say, choice measurement instruments, mode of administration, or sampling frame do not cause the data to be effectively divided into subsets that are simply not comparable?

Surveys often use mixed modes of administration, e.g. mail and telephone, and while this can improve response rates, the mode of administration can affect results \cite{Brambilla1987,McHorney1994} and differences in response behavior can lead to biased results. Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful.

The rapid growth of web surveys, due to low cost, timeliness, and other factors, generate large data sources that lack a sampling frame of the general population. However, it can be problematic to combine online panels (pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions) \cite{Liu2016}.

@@

\subsection{Principal component analysis}

Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^d$ denote the empirical covariance.

\emph{Principal component analysis} (PCA) may be interpreted in terms of the \emph{rank-q-reconstruction error}. Suppose that we want to describe each of the observations by $q$ numbers instead of the original $d$ numbers. The rank-q-reconstruction error is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*}
PCA ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of $S$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different. If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{j=q+1}^d \lambda_j$. The eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}. The $j$th loading can also found as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}
It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$. This implies that the loadings and variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension $q$.

@@

\vspace{\fill}\clearpage
\newpage
\bibliographystyle{apa}
\bibliography{bib}

\end{document}
